\documentclass[12pt, leqno]{article}
\input{common}

\begin{document}
\hdr{2020-01-31}

\section*{Binary floating point}

Why do we study conditioning of problems?  One reason is that we may have
input data contaminated by noise, resulting in a bad solution even if the
intermediate computations are done perfectly accurately.  But our concern in
this class is less with measurement errors and more with numerics, and so
we will study condition numbers with an eye to floating point error analysis.

Binary floating point arithmetic is essentially scientific notation.
Where in decimal scientific notation we write
\[
  \frac{1}{3}
  = 3.333\ldots \times 10^{-1},
\]
in floating point, we write
\[
  \frac{(1)_2}{(11)_2}
  = (1.010101\ldots)_2 \times 2^{-2}.
\]
Because computers are finite, however, we can only keep a finite
number of bits after the binary point.

In general, a {\em normal floating point number} has the form
\[
  (-1)^s \times (1.b_1 b_2 \ldots b_p)_2 \times 2^{E},
\]
where $s \in \{0,1\}$ is the {\em sign bit},
$E$ is the {\em exponent}, and
$(1.b_2 \ldots b_p)_2$ is the {\em significand}.
In the 64-bit double precision format, $p = 52$ bits are used to store
the significand, 11 bits are used for the exponent, and one bit is
used for the sign.  The valid exponent range for normal floating point
numbers is $-1023 < E < 1024$; this leaves two exponent encodings left
over for special purpose.  One of these special exponents is used to
encode {\em subnormal numbers} of the form
\[
  (-1)^s \times (0.b_1 b_2 \ldots b_p)_2 \times 2^{-1022};
\]
the other special exponent is used to encode $\pm \infty$ and NaN
(Not a Number).

For a general real number $x$, we will write
\[
  \fl(x) = \mbox{ correctly rounded floating point representation of } x.
\]
By default, ``correctly rounded'' means that we find the closest
floating point number to $x$, breaking any ties by rounding to the
number with a zero in the last bit\footnote{%
There are other rounding modes beside the default, but we will not
discuss them in this class}.
If $x$ exceeds the largest normal floating point number,
then $\fl(x) = \infty$.

\section*{Basic floating point arithmetic}

For basic operations (addition, subtraction, multiplication,
division, and square root), the floating point standard specifies that
the computer should produce the {\em true result, correctly rounded}.
So the MATLAB statement
\lstset{language=matlab,columns=flexible}
\begin{lstlisting}
    % Compute the sum of x and y (assuming they are exact)
    z = x + y;
\end{lstlisting}
actually computes the quantity $\hat{z} = \fl(x+y)$.  If $\hat{z}$ is
a normal double-precision floating point number, it will agree with
the true $z$ to 52 bits after the binary point.  That is, the relative
error will be smaller in magnitude than the {\em machine epsilon}
$\macheps = 2^{-53} \approx 1.1 \times 10^{-16}$:
\[
  \hat{z} = z(1+\delta), \quad |\delta| < \macheps.
\]
More generally, basic operations that produce normalized numbers are
correct to within a relative error of $\macheps$.  The floating
point standard also recommends that common transcendental functions,
such as exponential and trig functions, should be correctly rounded,
though compliant implementations that do not comply with this
recommendation may produce results with a relative error
just slightly larger than $\macheps$.

The fact that normal floating point results have a relative error
less than $\macheps$ gives us a useful {\em model} for reasoning about
floating point error.  We will refer to this as the ``$1 + \delta$''
model.  For example, suppose $x$ is an exactly-represented input to
the MATLAB statement
\begin{lstlisting}
    z = 1-x*x;
\end{lstlisting}
We can reason about the error in the computed $\hat{z}$ as follows:
\begin{align*}
  t_1 &= \fl(x^2) = x^2 (1+\delta_1) \\
  t_2 &= 1-t_1 = (1-x^2)\left( 1 + \frac{\delta_1 x^2}{1-x^2} \right) \\
  \hat{z}
  &= \fl(1-t_1)
    = z \left( 1 + \frac{\delta_1 x^2}{1-x^2} \right)(1+\delta_2) \\
  & \approx z \left( 1 + \frac{\delta_1 x^2}{1-x^2} +\delta_2 \right),
\end{align*}
where $|\delta_1|, |\delta_2| \leq \macheps$.  As before, we throw
away the (tiny) term involving $\delta_1 \delta_2$.
Note that if $z$ is close to zero (i.e.~if there is {\em cancellation} in the
subtraction), then the model shows the result may have a
large relative error.

\section*{Exceptions}

We say there is an {\em exception} when the floating point result is
not an ordinary value that represents the exact result.  The most
common exception is {\em inexact} (i.e. some rounding was needed).
Other exceptions occur when we fail to produce a normalized floating
point number.  These exceptions are:
\begin{description}
\item[Underflow:]
  An expression is too small to be represented as a normalized floating
  point value.  The default behavior is to return a subnormal.
\item[Overflow:]
  An expression is too large to be represented as a floating point
  number.  The default behavior is to return {\tt inf}.
\item[Invalid:]
  An expression evaluates to Not-a-Number (such as $0/0$)
\item[Divide by zero:]
  An expression evaluates ``exactly'' to an infinite value
  (such as $1/0$ or $\log(0)$).
\end{description}
When exceptions other than inexact occur, the usual ``$1 + \delta$''
model used for most rounding error analysis is not valid.

\section*{Finding and fixing floating point problems}

Floating point arithmetic is not the same as real arithmetic.  Even
simple properties like associativity or distributivity of addition and
multiplication only hold approximately.  Thus, some computations that
look fine in exact arithmetic can produce bad answers in floating
point.  What follows is a (very incomplete) list of some of the ways
in which programmers can go awry with careless floating point
programming.

\subsection*{Cancellation}

If $\hat{x} = x(1+\delta_1)$ and $\hat{y} = y(1+\delta_2)$ are
floating point approximations to $x$ and $y$ that are very close, then
$\fl(\hat{x}-\hat{y})$ may be a poor approximation to $x-y$ due to
{\em cancellation}.  In some ways, the subtraction is blameless in
this tail: if $x$ and $y$ are close, then $\fl(\hat{x}-\hat{y}) =
\hat{x}-\hat{y}$, and the subtraction causes no additional rounding
error.  Rather, the problem is with the approximation error already
present in $\hat{x}$ and $\hat{y}$.

The standard example of loss of accuracy revealed through cancellation
is in the computation of the smaller root of a quadratic using the
quadratic formula, e.g.
\[
  x = 1-\sqrt{1-z}
\]
for $z$ small.  Fortunately, some algebraic manipulation gives an
equivalent formula that does not suffer cancellation:
\[
  x =
    \left( 1-\sqrt{1-z} \right)
    \left(\frac{1+\sqrt{1-z}}{1+\sqrt{1-z}}\right)
  =
    \frac{z}{1+\sqrt{1-z}}.
\]

\subsection*{Sensitive subproblems}

We often solve problems by breaking them into simpler subproblems.
Unfortunately, it is easy to produce badly-conditioned subproblems
as steps to solving a well-conditioned problem.  As a simple (if
contrived) example, try running the following MATLAB code:

\lstset{language=matlab,frame=lines,columns=flexible}
\lstinputlisting{code/sillysqrt.m}

In exact arithmetic, this should produce 2, but what does it produce
in floating point?  In fact, the first loop produces a correctly
rounded result, but the second loop represents the function
$x^{2^{60}}$, which has a condition number far greater than $10^{16}$
--- and so all accuracy is lost.

\subsection*{Unstable recurrences}

We gave an example of this problem in the last lecture notes when we
looked at the recurrence
\begin{align*}
  E_0 &= 1-1/e \\
  E_n &= 1-nE_{n-1}, \quad n \geq 1.
\end{align*}
No single step of this recurrence causes the error to explode, but
each step amplifies the error somewhat, resulting in an exponential
growth in error.

\subsection*{Undetected underflow}

In Bayesian statistics, one sometimes computes ratios of long
products.  These products may underflow individually, even when the
final ratio is not far from one.  In the best case, the products will
grow so tiny that they underflow to zero, and the user may notice an
infinity or NaN in the final result.  In the worst case, the
underflowed results will produce nonzero subnormal numbers with
unexpectedly poor relative accuracy, and the final result will be
wildly inaccurate with no warning except for the (often ignored)
underflow flag.

\subsection*{Bad branches}

A NaN result is often a blessing in disguise: if you see an
unexpected NaN, at least you {\em know} something has gone wrong!
But all comparisons involving NaN are false,  and so
when a floating point result is used to compute a branch condition
and an unexpected NaN appears, the result can wreak havoc.
As an example, try out the following code in MATLAB.

\lstset{language=matlab,frame=lines,columns=flexible}
\lstinputlisting{code/testnan.m}

\section*{Limits of the ``$1+\delta$'' model}

Apart from the fact that it fails when there are exceptions other than
inexact, the ``$1+\delta$'' model of floating point does not reflect
the fact that some computations involve no rounding error.  For example:
\begin{itemize}
\item
  If $x$ and $y$ are floating point numbers within a factor of two of
  each other, $\fl(x-y)$ is computed without rounding error.
\item
  Barring overflow or underflow to zero, $\fl(2x) = 2x$ and $\fl(x/2) = x/2$.
\item
  Integers between $\pm (2^{53}-1)$ are represented exactly.
\end{itemize}
These properties of floating point allow us to do some clever things,
such as using ordinary double precision arithmetic to simulate
arithmetic with about twice the number of digits.  You should be aware
that these tricks exist, even if you never need to implement them --
otherwise, I may find myself cursing a compiler you
wrote for rearranging the computations in a floating point code that I
wrote!

Let's consider an example analysis that illustrates both the
usefulness and the limitations of the $1+\delta$ model.  Suppose
I have points $A$, $B$, and $C$ in the box $[1,2]^2$, and that the
coordinates are given as exact floating point numbers.  I want to
determine whether $C$ is above, below, or on the oriented line going through
$A$ and $B$.  The usual way to do this is by looking at the sign of
\[
  \det \begin{bmatrix} B_x-A_x & C_x-A_x \\ B_y-A_y & C_y-A_y \end{bmatrix}.
\]

We might compute this determinant as something like this:
\begin{align*}
  t_1 &= B_x-A_x \\
  t_2 &= B_y-A_y \\
  t_3 &= C_x-A_x \\
  t_4 &= C_y-A_y \\
  t_5 &= t_1 \times t_4 \\
  t_6 &= t_2 \times t_3 \\
  t_7 &= t_5-t_6.
\end{align*}
Now, suppose we do this computation in floating point.  We will call
the floating point results $\hat{t}_j$, to distinguish them from the
exact results.  According to the $1+\delta$ model,
we have $|\delta_i| \leq \epsilon$ so that
\begin{align*}
  \hat{t}_1 &= (B_x-A_x)(1+\delta_1) \\
  \hat{t}_2 &= (B_y-A_y)(1+\delta_2) \\
  \hat{t}_3 &= (C_x-A_x)(1+\delta_3) \\
  \hat{t}_4 &= (C_y-A_y)(1+\delta_4) \\
  \hat{t}_5 &= (\hat{t}_1 \times \hat{t}_4)(1+\delta_5)
             = (t_1 \times t_4)(1+\delta_1)(1+\delta_4)(1+\delta_5)
             = t_5 (1+\gamma_5) \\
  \hat{t}_6 &= (\hat{t}_2 \times \hat{t}_3)(1+\delta_6)
             = (t_2 \times t_3)(1+\delta_2)(1+\delta_3)(1+\delta_6)
             = t_6 (1+\gamma_6) \\
  \hat{t}_7 &= (\hat{t}_5-\hat{t}_6)(1+\delta_7) \\
            &= (t_5-t_6)
               \left( 1 + \frac{t_5 \gamma_5 - t_6 \gamma_6}{t_5-t_6} \right)
               (1 + \delta_7).
\end{align*}
Here, $1 + \gamma_5 = (1+\delta_1)(1+\delta_2)(1+\delta_3)
= 1+\delta_1 + \delta_2 + \delta_3 + O(\epsilon^2)$; that is,
$|\gamma_5| \lesssim 3 \epsilon$.
Similarly, $|\gamma_6| \lesssim 3_{\epsilon}$.

Now, how large can the relative error in $\hat{t}_7$ be?  Ignoring the final
rounding error $\delta_7$ -- which is in this case insignficant -- we have
that the relative error is bounded by
\[
  \left| \frac{t_5 \gamma_5 - t_6 \gamma_6}{t_5-t_6} \right| \leq
  3 \epsilon \frac{|t_5| + |t_6|}{|t_5-t_6|}.
\]
If $t_5$ and $t_6$ are not small but $t_5-t_6$ is small, the relative error
in $t_7$ could be quite large -- even though the absolute error remains small.
This effect of a large relative error due to a small result in a subtraction
is called {\em cancellation}.  In this case,
if the relative error is one or larger, then we don't even necessarily
have the right sign!  That's not a good thing for our test.

Is this error analysis pessimistic?  Yes and no.  Some things in
floating point are {\em exact}, such as multiplication by a power of
two or subtraction of two numbers within a factor of two of each
other.  Thus, there will be no rounding error in the first four steps
($\delta_i = 0$ for $i = 1, 2, 3, 4$), since we have constrained each
of the coordinates to lie in the interval $[1,2]$.  Note that this
means that if $t_5$ and $t_6$ are close to each other, there will be
no rounding error in the last step -- $\delta_7$ will be zero!  But
the overall outlook remains grim: the rounding errors in the
multiplications are generally really there; and even if the
subtraction is done exactly, it is the propogated error from the
multiplication steps that destroys the relative accuracy of the final
results.

For this computation, then, the critical rounding error is really in
the multiplications.  If we could do those multiplications exactly,
then we would be able to compute the desired determinant to high relative
accuracy (and thus reliably test the determinant's sign).  As it turns
out, we {\em can} perform the multiplication exactly if we use a higher
precision for intermediate computations than for the input data.
For example, suppose the input data are in single precision, but the
intermediate computations are all performed in double precision.
Then  $t_1$ through $t_4$ can be represented with significands that are at most
24 bits, and the products $t_5$ and $t_6$ require at most 48 bits --
and can thus be exactly represented as double precision numbers with
53 bits in the significand.

For this computation, we see a payoff by doing a more detailed analysis
than is possible with only the $1+\delta$ model.  But the details
were tedious, and analysis with the $1+\delta$ model will usually
be good enough.

\section*{Problems to ponder}

\begin{enumerate}
\item
  How do we accurately evaluate $\sqrt{1+x}-\sqrt{1-x}$ when $x \ll 1$?
\item
  How do we accurately evaluate $\ln \sqrt{x+1}-\ln\sqrt{x}$ when $x \gg 1$?
\item
  How do we accurately evaluate $(1-\cos(x))/\sin(x)$ when $x \ll 1$?
\item
  How would we compute $\cos(x)-1$ accurately when $x \ll 1$?
\item
  The {\em Lamb-Oseen vortex} is a solution to the 2D Navier-Stokes
  equation that plays a key role in some methods for computational
  fluid dynamics.  It has the form
  \[
    v_{\theta}(r,t) = \frac{\Gamma}{2 \pi r}
      \left( 1 - \exp\left( \frac{-r^2}{4 \nu t} \right) \right)
  \]
  How would one evaluate $v(r,t)$ to high relative accuracy for all
  values of $r$ and $t$ (barring overflow or underflow)?
\item
  For $x > 1$, the equation $x = \cosh(y)$ can be solved as
  \[
    y = -\ln\left(x - \sqrt{x^2-1}\right).
  \]
  What happens when $x = 10^8$?  Can we fix it?
\item
  The difference equation
  \[
    x_{k+1} = 2.25x_k - 0.5 x_{k-1}
  \]
  with starting values
  \[
    x_1 = \frac{1}{3}, \qquad x_2 = \frac{1}{12}
  \]
  has solution
  \[
    x_k = \frac{4^{1-k}}{3}.
  \]
  Is this what you actually see if you compute?  What goes wrong?
\item
  Considering the following two MATLAB fragments:
\begin{lstlisting}
    % Version 1
    f = (exp(x)-1)/x;

    % Version 2
    y = exp(x);
    f = (1-y)/log(y);
\end{lstlisting}
  In exact arithmetic, the two fragments are equivalent.  In floating
  point, the first formulation is inaccurate for $x \ll 1$, while the
  second formulation remains accurate.  Why?
\item
  Running the recurrence $E_n = 1-nE_{n-1}$ {\em forward} is an
  unstable way to compute $\int_0^1 x^n e^{x-1} \, dx$.  However, we
  can get good results by running the recurrence {\em backward} from
  the estimate $E_n \approx 1/(N+1)$ starting at large enough $N$.
  Explain why.  How large must $N$ be to compute $E_{20}$ to near
  machine precision?
\item
  How might you accurately compute this function for $|x| < 1$?
  \[
    f(x) = \sum_{j=0}^\infty \left( \cos(x^j)-1 \right)
  \]
\end{enumerate}

\end{document}
